---
title: "FML Assignment 2"
author: "Pooja Babu"
date: "2025-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the required packages

```{r cars}
library(caret)
library(ISLR)
```

## View the Carseats dataset

```{r}
head(Carseats,n=10)
```

## Checking the dimensions of the dataset

```{r}
dim(Carseats)
```

## Summary of the dataset

```{r}
summary(Carseats)
```

## Splitting the dataset into 3 sets (training_data, validation_data, test_data)

### Step 1: Partitioning carseats datasets into temp (300 records) and test (100 records)
```{r}
temp = createDataPartition(y=1:nrow(Carseats),p=300/nrow(Carseats),list = FALSE)
```
### Checking the dimensions of temp dataset
```{r}
dim(temp)
```

### Subsetting the dataset using temp so that the values in train_index (300 records) does not repeat in test_data (100 records)
```{r}
train_index=Carseats[temp,]
```


```{r}
test_data=Carseats[-temp,]
```
### Checking the dimensions of train_index and test_data
```{r}
dim(train_index)
```
```{r}
dim(test_data)
```
## Step 2: Now we divide the train_index into 2 sets in such a way that 200 records are put in training_data and 100 records in validation_data.

### We create a temp_train_index to store the 200 records that we select randomly from the train_index

```{r}
temp_train_index<-createDataPartition(y=1:nrow(train_index),p=200/nrow(train_index),list = FALSE)
```
### Checking the dimensions of temp_train_index
```{r}
dim(temp_train_index)
```
### Subsetting the data into training_data (200 records) and validation_data(100 records)
```{r}
training_data=train_index[temp_train_index,]
```


```{r}
validation_data=train_index[-temp_train_index,]
```
### Checking the dimensions of training_data and validation_data
```{r}
dim(training_data)
```

```{r}
dim(validation_data)
```
## Checking the first 6 rows of all the datasets
```{r}
head(validation_data)
```


```{r}
head(training_data)
```


```{r}
head(test_data)
```
## Summary of training_data , validation_data and test_data
```{r}
summary(training_data)
```

```{r}
summary(validation_data)
```

```{r}
summary(test_data)
```



